Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-05-21 manuscript/manuscript.Rmd
updates figures in ms

```{r libraris, include=FALSE}
library(ggplot2)
library(plyr)
library(gridExtra)
library(reshape2)
library(xtable)
```

@@

stats_table <- rbind(ipmstats, qbmstats)
stats_table <- stats_table[ order(stats_table[,"species"]), ]
stats_table <- stats_table[,c("model", "species", "mae",
                              "rho","mean_cover")]

colnames(stats_table) <- c("Species", "Model", "MAE", "Accuracy", "Mean Obs. Cover")
print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation."),type="latex", comment=FALSE, include.rownames=FALSE)
```

@@

>>>>

```{r libraris, include=FALSE}
library(ggplot2)
library(plyr)
library(gridExtra)
library(reshape2)
library(xtable)
library(ggmcmc)
```

@@

stats_table <- rbind(ipmstats, qbmstats)
stats_table <- stats_table[ order(stats_table[,"species"]), ]
stats_table <- stats_table[,c("species", "model", "mae",
                              "rho","mean_cover")]
# comment          <- list()
# comment$pos      <- list()
# comment$pos[[1]] <- c(nrow(stats_table))
# comment$command  <- c(paste("\\hline \n",  # we`ll replace all default hlines with this and the ones below
#                             "The IPM MAE is significantly lower for HECO ($p = 3.3\times10^{-10}$) and POSE ($p = 0.00013$. MAEs are statisticially similar between models for BOGR and PASM.   \n",
#                             sep = ""))

colnames(stats_table) <- c("Species", "Model", "MAE", "Accuracy", "Mean Obs. Cover")
print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation."),type="latex", comment=FALSE,
             include.rownames=FALSE, caption.placement="top")
```
The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$) and POSE ($p = 0.00013$). MAEs are statisticially similar between models for BOGR and PASM.

@@

```{r ipm_climeffs, cache=FALSE, results='hide', include=FALSE}
source("../montana/ipm/vitalRateRegs/climate_effects_plots.R")
```

![Standardized regression coefficients for climate effects in the vital rate models for the IPM. Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines). Left panels (black) are for the growth regressions; middle panels (blue) are for the survival regressions; right panels (orange) are for the recruitment regressions.](components/figure/ipm_climeffs.pdf)

```{r quad_climeffs, cache=FALSE, results='hide', include=FALSE}
source("../montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R")
```

![Standardized regression coefficients for climate effects in the vital rate models for the QBM. Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).](components/figure/qbm_climeffs.pdf)

### 2015-05-21 manuscript/manuscript.tex
updates figures in ms

\documentclass[12pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
    \setmainfont{times}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\author{}
\date{}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing

@@

\normalsize

@@

\emph{Andrew T. Tredennick
(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, UT}

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
dataset is freely available on Ecological Archives (Anderson et al.
2011), and interested readers should refer to the metadata therein for a
complete description. The site is about 800 m above sea level and mean
annual precipitation (1878-2009) is 334 mm, with most annual
precipitation falling from April through September. The site is grass
dominated and, for the purposes of our study, we focus on the four most
abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annually
in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing$\times$climate interactions. The annual maps of the
quadrats were digitized and the fates of individual plants tracked and
extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

@@

We fit all models using a hierarchical Bayesian approach, which we
describe in more detail below. However, for each vital rate statistical
model we also define the likelihood model we use. For the likelihood
models, \textbf{Y} is always the relevant vector of observations (e.g.,
whether a genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).

@@

We used logistic regression to model survival probability ($S$) of genet
$i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:

@@

where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a
year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope
parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat
group location, and $\theta^{S}_{k}$ is the fixed parameter for the
effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We
include density-dependence by estimating the effect of crowding on the
focal individual by other individuals of the same species. $\omega$ is
the effect of crowding and $w_{t,Q}$ is the crowding experienced by the
focal individual at time $t$ in quadrat group $Q$.

We modeled growth as Gaussian process describing genet size at time
$t+1$ as a function of size at $t$ and climate covariates:

@@

where $x$ is log genet size and all other parameters are as described
for the survival regression.

Our data allows us to track new recruits, but we cannot assign a
specific parent to new genets. So, for recruitment, we work at the
quadrat level and model the number of new individuals of species $j$ in
quadrat $q$ recruiting at time $t+1$ as a function of quadrat
``effective cover'' ($A'$) in the previous year ($t$). Effective cover
is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which
$q$ is located:

@@

where $p$ is a mixing fraction between 0 and 1 that is estimated within
the model.

We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$
follows a negative binomial distribution:

@@

where $\lambda$ is the mean intensity and $\zeta$ is the size parameter.
We define $\lambda$ as:

@@

where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat
$q$ and all other terms are as in the survival and growth regressions.

@@

The statistical approach used to model vital rates using aggregated data
depends on the type of data collected. In our case, and as is often the
case with census data, we have percent cover data (which can easily be
transformed to proportion data, of course). We first considered fitting
three vital rate models analagous to those we fit at the individual
level: one for probability of extirpation within a quadrat (analagous to
survival), one for cover change within a quadrat (analagous to growth),
and one for probability of colonization within a quadrat (analagous to
recruitment). However, within-quadrat extirpation and colonization
events were rare in our time series ($N=9$ and $N=10$, respectively
across all species). Given the broad spatial distribution of the
quadrats we are studying, it is safe to assume that these events are in
fact rare enough to be ignored for our purposes. So we constrained our
statistical modeling of vital rates at the population level to change in
percent cover within quadrats. For the remaining discussion of
statistical modeling we refer to proportion data, which is simply
percent data divided by 100.

An obvious choice for fitting a linear model to proportion data is beta
regression because the support of the beta distribution is {[}0,1{]},
not including true zeros or ones. However, when we used fitted model
parameters from a beta regression in a quadrat-based population model
the simulated population tended toward 100\% cover for all species. We
therefore chose a more constrained modeling approach based on a
truncated log-normal likelihood. The model for quadrat cover change
($G$) from time $t$ to $t+1$ is

@@

where $x_{jq,t}$ is the log of species' $j$ proportional cover in
quadrat $q$ at time $t$ and all other parameters are as in the
individual-level growth model (Eq. \#). The log normal likelihood
includes a truncation ($T${[}0,1{]}) to ensure that predicted values do
not exceed 100\% cover.

@@

where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th
climate covariate.

@@

\begin{table}[ht]
\centering
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & Accuracy & Mean Obs. Cover \\ 
  \hline
IPM & BOGR & 4.77 & 0.90 & 9.16 \\ 
  QBM & BOGR & 5.30 & 0.88 & 9.22 \\ 
  IPM & HECO & 0.60 & 0.80 & 1.22 \\ 
  QBM & HECO & 2.03 & 0.84 & 1.25 \\ 
  IPM & PASM & 0.18 & 0.65 & 0.40 \\ 
  QBM & PASM & 0.17 & 0.87 & 0.40 \\ 
  IPM & POSE & 0.78 & 0.61 & 1.23 \\ 
  QBM & POSE & 1.30 & 0.57 & 1.25 \\ 
   \hline
\end{tabular}
\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.} 
\end{table}

@@

@@

\subsection{References}\label{references}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
    \setmainfont{times}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}
\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing

@@

\maketitle

@@

\emph{Andrew T. Tredennick
(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}),
Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, UT}

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W).
The dataset is freely available on Ecological Archives (Anderson et al.
2011), and interested readers should refer to the metadata therein for a
complete description. The site is about 800 m above sea level and mean
annual precipitation (1878-2009) is 334 mm, with most annual
precipitation falling from April through September. The site is grass
dominated and, for the purposes of our study, we focus on the four most
abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annually
in 44 1-\(\text{m}^2\) quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing\(\times\)climate interactions. The annual maps of
the quadrats were digitized and the fates of individual plants tracked
and extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

@@

We fit all models using a hierarchical Bayesian approach, which we
describe in more detail below. However, for each vital rate statistical
model we also define the likelihood model we use. For the likelihood
models, \textbf{Y} is always the relevant vector of observations (e.g.,
whether a genet survived {[}1{]} or not {[}0{]} from year \(t\) to
\(t+1\)).

@@

We used logistic regression to model survival probability (\(S\)) of
genet \(i\) from species \(j\) in quadrat group \(Q\) from time \(t\) to
\(t+1\):

@@

where \(x_{ij,t}\) is the log of genet size, \(\gamma^{S}_{j,t}\) is a
year-specific intercept, \(\beta^{S}_{j,t}\) is the year-specific slope
parameter for size, \(\phi^{S}_{jQ}\) is the random effect of quadrat
group location, and \(\theta^{S}_{k}\) is the fixed parameter for the
effect of the \(k\)th climate covariate at time \(t\) (\(C_{k,t}\)). We
include density-dependence by estimating the effect of crowding on the
focal individual by other individuals of the same species. \(\omega\) is
the effect of crowding and \(w_{t,Q}\) is the crowding experienced by
the focal individual at time \(t\) in quadrat group \(Q\).

We modeled growth as Gaussian process describing genet size at time
\(t+1\) as a function of size at \(t\) and climate covariates:

@@

where \(x\) is log genet size and all other parameters are as described
for the survival regression.

Our data allows us to track new recruits, but we cannot assign a
specific parent to new genets. So, for recruitment, we work at the
quadrat level and model the number of new individuals of species \(j\)
in quadrat \(q\) recruiting at time \(t+1\) as a function of quadrat
``effective cover'' (\(A'\)) in the previous year (\(t\)). Effective
cover is a mixture of observed cover (\(A\)) in the focal quadrat
(\(q\)) and the mean cover across the entire group (\(\bar{A}\)) of
\(Q\) quadrats in which \(q\) is located:

@@

where \(p\) is a mixing fraction between 0 and 1 that is estimated
within the model.

We assume the number of individuals, \(Y^{R}\), recruiting at time
\(t+1\) follows a negative binomial distribution:

@@

where \(\lambda\) is the mean intensity and \(\zeta\) is the size
parameter. We define \(\lambda\) as:

@@

where \(A'\) is effective cover (\(\text{cm}^2\)) of species \(j\) in
quadrat \(q\) and all other terms are as in the survival and growth
regressions.

@@

The statistical approach used to model vital rates using aggregated data
depends on the type of data collected. In our case, and as is often the
case with census data, we have percent cover data (which can easily be
transformed to proportion data, of course). We first considered fitting
three vital rate models analagous to those we fit at the individual
level: one for probability of extirpation within a quadrat (analagous to
survival), one for cover change within a quadrat (analagous to growth),
and one for probability of colonization within a quadrat (analagous to
recruitment). However, within-quadrat extirpation and colonization
events were rare in our time series (\(N=9\) and \(N=10\), respectively
across all species). Given the broad spatial distribution of the
quadrats we are studying, it is safe to assume that these events are in
fact rare enough to be ignored for our purposes. So we constrained our
statistical modeling of vital rates at the population level to change in
percent cover within quadrats. For the remaining discussion of
statistical modeling we refer to proportion data, which is simply
percent data divided by 100.

An obvious choice for fitting a linear model to proportion data is beta
regression because the support of the beta distribution is {[}0,1{]},
not including true zeros or ones. However, when we used fitted model
parameters from a beta regression in a quadrat-based population model
the simulated population tended toward 100\% cover for all species. We
therefore chose a more constrained modeling approach based on a
truncated log-normal likelihood. The model for quadrat cover change
(\(G\)) from time \(t\) to \(t+1\) is

@@

where \(x_{jq,t}\) is the log of species' \(j\) proportional cover in
quadrat \(q\) at time \(t\) and all other parameters are as in the
individual-level growth model (Eq. \#). The log normal likelihood
includes a truncation (\(T\){[}0,1{]}) to ensure that predicted values
do not exceed 100\% cover.

@@

where \(\bar{\theta_{k}}\) is the interspecific effect of the \(k\)th
climate covariate.

@@

\begin{table}[ht]
\centering
\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & Accuracy & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 4.77 & 0.90 & 9.16 \\ 
  BOGR & QBM & 5.30 & 0.88 & 9.22 \\ 
  HECO & IPM & 0.60 & 0.80 & 1.22 \\ 
  HECO & QBM & 2.03 & 0.84 & 1.25 \\ 
  PASM & IPM & 0.18 & 0.65 & 0.40 \\ 
  PASM & QBM & 0.17 & 0.87 & 0.40 \\ 
  POSE & IPM & 0.78 & 0.61 & 1.23 \\ 
  POSE & QBM & 1.30 & 0.57 & 1.25 \\ 
   \hline
\end{tabular}
\end{table}

The IPM MAE is significantly lower for HECO
(\(p = 3.3 \times 10^{-10}\)) and POSE (\(p = 0.00013\)). MAEs are
statisticially similar between models for BOGR and PASM

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/ipm_climeffs.pdf}
\caption{Standardized regression coefficients for climate effects in the
vital rate models for the IPM. Shown are the median (circles), 75\%
confidence intervals (thick lines), and 95\% confidence intervals (thin
lines). Left panels (black) are for the growth regressions; middle
panels (blue) are for the survival regressions; right panels (orange)
are for the recruitment regressions.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/qbm_climeffs.pdf}
\caption{Standardized regression coefficients for climate effects in the
vital rate models for the QBM. Shown are the median (circles), 75\%
confidence intervals (thick lines), and 95\% confidence intervals (thin
lines).}
\end{figure}

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}
